Quantifying Stability and Change in Personal Culture Using Panel Data

### 2023-09-27 manuscript/manuscript.tex
cleaning up and updates

A recurring focus in the sociology of culture is the question of whether
people, as they move through their lives and transition between social
roles, change their personal culture --- the declarative and
non-declarative attitudes, beliefs, values, practices, and dispositions
that exist at the individual level (Kiley and Vaisey 2020; Lersch 2023;
Lizardo 2017).

This seemingly simple question underlies a number of debates across
sociological subfields. For example, pragmatist theories of action claim
that changes in social environments cause people to adapt their views
and make new cultural meanings (Gross 2009; Swidler 2001), while
Bourdieusian practice theories argue that the ``past conditions of
production'' leave a mark on people's personal culture that lasts
throughout their lives (Bourdieu 1990). Similarly, models of social
influence assume that people adapt their culture in the face of new
information (Vaisey and Lizardo 2010 KK Note: I don't feel like this is
a good cite for this.), while the emphasis on cohort effects in models
of aggregate social and cultural change requires them to be open to
change while young but become fairly resistant to it as they age (Ryder
1965). Finally, life course theories posit slower but still important
changes over time as people advance through important transitions in
their lives (Bardi et al. 2009; Elder, Johnson, and Crosnoe 2003).

Because all of the above theoretical perspectives have some empirical
support, the current debate revolves around not whether they exist, but
when and how they each apply and the contribution of each to explaining
cultural differences. That being said, scholars have struggled to reach
a consensus on the relative importance of change during adulthood. Over
the span of a few years, most apparent changes in people's responses to
personal culture items appear to be transitory, with little evidence of
persistent change among adults (Kiley and Vaisey 2020; Vaisey and Kiley
2021). This suggests that change during adulthood is not a strong force
in explaining personal cultural differences. When looked at over longer
time horizons, however, there is at least some evidence that adults make
persistent cultural changes, especially as they move through distinct
social roles (Lersch 2023).

These seemingly inconsistent findings are partly due to the fact that
researchers have taken a ``tournament of models'' approach in
adjudicating different theoretical perspectives (Lersch 2023: p.~228).
Accordingly, researchers have formalized models of responses to cultural
items over time as demonstrating either ``change'' or ``no change,''
which essentially asks whether we \emph{ever} observe intrapersonal
change in personal culture.

We argue that asking only \emph{whether} people change obscures what are
actually highly compatible empirical patterns. Being able to detect
change in a given dataset depends not only on sample populations, item
measurement, statistical methods, and power, but crucially also on
definitions of what counts as ``change.'' These methodological decisions
potentially limit the possibility of convergence on the theoretical
debates. Moreover, asking whether there is evidence of change limits the
kind of questions that can be asked to, ``Do we see change in this
measure?'' But such inquiries are ill-suited to address what's at the
core of the theoretical debate. There is enough evidence to assume that
neither socialization at an early age nor the influence of the
contemporary environment alone can explain cultural differences. The
question that remains is under what conditions, across which groups, and
in what domains do we observe differences in the relative importance of
intrapersonal change and interpersonal differences. This focus is
necessarily comparative and requires an approach that quantifies forms
of cultural change rather than calling victory for the one or the other
perspective.

In this paper, we try to intervene in this debate by moving beyond the
``tournament of models'' approach toward a quantification of the
processes that lead to differences in personal culture. Drawing on the
seven panel datasets analyzed in the debate, we use established methods
to quantify the amount of variance in responses that is due to
between-people differences at a single point during the survey versus
the amount of variance in responses that is due to linear intrapersonal
change over time. This measure gives us a sense of the importance of
change over the duration of the panel relative to the differences that
exist between people assuming they never changed.

@@

Despite differences in how these researchers interpret their findings,
the empirical results are not far apart. Kiley and Vaisey (2020) found
that the majority of items they tested favored the AUM, meaning there
was evidence of durable change among adults, a pattern consistent with
Lersch's results. While Lersch (2023) found evidence of durable change
on most items, this change only accounted for a small proportion of
variance explained. On average, people changed only about .07 standard
deviations over 10 years. This is consistent with Kiley and Vaisey's
finding that durable change is often small. And a quarter of the items
Lersch studied still preferred the SDM, suggesting that even with
different assumptions assumptions about change and more extensive data,
many questions show no evidence of durable change. Likewise, in studies
showing evidence of durable change in specific measures of personal
culture, the change is often small. For example, Brocic and Miles (2021)
estimate that completing graduate degrees in humanities, arts, and
social sciences shift peoples' moral relativism only about .2 standard
deviations on average compared to people with no college degree, and
this is the largest effect they identify. And studies of aggregate
change show that, even on items where cohort effects explain more
variance than period effects, there is always evidence that some people
change over time (Vaisey and Lizardo 2016). In other words, despite
different interpretations, the results of previous work are in many ways
highly consonant. {[}I think the revisions that were done to this
paragraph were confusing and mostly reverted it.{]}

@@

A more theoretically productive approach begins with a model assuming
that during an observed time period people might remain stable, might
change a little, or might undergo significant shifts in their personal
culture. Lersch's LCAM formally does this by modeling each individual as
following their own trajectory over time. But rather than debating
whether this model fits a particular data set better than a model that
assumes that people never change, researchers should quantify individual
changes over time and compare those to the differences that exist
between people at the panel's start. In other words, the focus should
shift from asking, ``do people change?'' to asking, ``what are the
relative contributions of change and initial differences toward
explaining differences in personal culture?'' {[}KK Note: I didn't like
this change. It's not about collectively explaining, its about them
relative to each other.{]}

In this approach, there are two metrics that each reflect a combination
of relevant theoretical processes. The first metric, interpersonal
differences at the start of the survey, reflects the accumulated
experiences of people prior to entering the panel survey that shape how
they answer questions, what Lersch calls ``early imprinting.'' While
commonly associated with experiences during an early formative period
that result in settled dispositions, this variation could also reflect
experiences that happen at any stage in life as long as they predate the
panel and consistently impact subsequent responses. For instance, for
those people entering the panel post-retirement, this ``imprinting''
might reflect this pivotal life transition. Consequently, this metric
also reflects variation in individuals' social roles or statuses at the
start of the panel that were important in shaping their dispositions.
{[}TK Note: I feel that this paragraph puts too much emphasis on
pre-panel processes. It might be early imprinting or broad life-course,
but why emphasize it this much? KK NOTE: Because it could be either, I
don't want to treat it like it's just ``imprinting.''{]} {[}KK NOTE: You
reversed what was being said here. You say ``early imprinting'' denotes
intrapersonal change (process denotes number). We want to say the
opposite -- what I had before -- number denotes process: ``interpersonal
differences'' denotes ``early imprinting.'' I changed it back. I also
took out reference to ``transformative'' experiences because I don't
think we're necessarily talking about those specifically, just any
experience that changes you. Because we don't say a specific theoretical
process (``socialization'', etc.) I want to keep the language of each
denoting combinations of processes. ``Persistent change'' isn't a
theoretical process. It's a bunch of them.{]}

@@

A third metric, often called ``residual variance'' or ``fluctuation,''
accounts for the remainder of variance in peoples' responses over time.
This non-durable change emerges for a variety of reasons. For example,
people might not have clear personal cultural dispositions as measured
by surveys. Instead, they might internalize a broad set of
considerations and construct an opinion in the context of the survey
interview, with different considerations coming to the forefront of
their cognition each interview (Feldman and Zaller 1992; Tourangeau,
Rips, and Rasinski 2000; Zaller 1992). This variance in considerations
can produce variance in responses even if peoples' ``dispositions''
--the full set of considerations they could draw on-- remains stable
over time. Additionally, this variance can also include measurement
errors, such as misinterpreted questions, erroneous response selections,
or responses getting coded incorrectly. While this third process
reflects important aspects of personal culture, it does not directly
touch on the ongoing debate outlined here. Because of this we quantify
this component but focus principally on the other two. {[}KK NOTE: I
agreed with TK's note and mostly reverted this paragraph.{]}

@@

Third, we expect the variance explained by each process --interpersonal
differences and intrapersonal change-- to depend on specific attributes
of the question and panel. First, the higher the response resolution for
a given question and the larger the sample, the greater the variance
explained by intrapersonal change should be, as smaller changes in
intrapersonal change will be easier to detect. Second, the LCAM and
various theoretical perspectives that underlie it posit that social
transitions drive changes in personal culture. Since the probability
that any person experiences such a transition goes up the longer we
observe them, we expect panel duration to be positively related to the
proportion of variation explained by intrapersonal change.

@@

\[
y_{i, t} = \beta_0 + u_{0, i} + \beta_{1} x_t + u_{1, i} x_{i, t} + \epsilon_{i, t}
\] \[
u_{0,i} \sim \mathcal{N}(0,\sigma_{\mu_0}^2)
\]

\[
u_{1,i} \sim \mathcal{N}(0,\sigma_{u_1}^2)
\]

{[}KKNOTE: Is that right? I forget how to do that exactly.{]}

@@

\[
\tau = \frac{V(C)}{V(C) + V(D)}
\]

{[}TK Note: \textbackslash phi taken in Kiley \& Vaisey,
\textbackslash delta taken in Vaisey \& Lizardo. I think using a Greek
letter without strong associations (something like alpha, beta, or mu or
whatever) is not good. So I settled on \textbackslash tau. Feel free to
change it.{]}

We can think of this number, \(\tau\) as akin to an Intraclass
Correlation Coefficient focused only on the systematic components of the
model while ignoring residual variance that might be attributable to
measurement error or non-persisting change (see also the variance
decomposition of period and cohort by Vaisey and Lizardo 2016). The
higher this number is, the more systematic variance in responses is
attributable to intrapersonal change during the panel. The lower the
number, the less intrapersonal change seems to play a large role in
explaining systematic cultural differences during the panel. Moving
forward, we use \(\tau\) as a placeholder for \emph{systematic variance
attributable to intrapersonal change}.

@@

As a second step in our analysis, we use ordinary least squares
regression to model \(\tau\) as a function of features of the question
and panel. This allows us to explore how these features affect the
prevalence of intrapersonal change. These covariates include the number
of response options (measured as 2, 3-5, and more than 5), the panel,
the log of the number of participants, the number of waves observed, the
time since the question was first asked, and the total span of time the
question was asked.

{[}TK Note: I guess we need to say something here about the college
stuff? KK Note: I didn't feel like that was necessary. It feels so
distinct later on that I think it's OK to kind of ``start over'' there.
I'm worried that people will forget the plan by the time they get there,
and we'll have to reiterate everything anyway. But others should weigh
in.{]}.

@@

Figure 1 plots the proportion of systematic variance attributable to
interpersonal differences and intrapersonal change for each of the 609
questions analyzed, plotted separately by panel. To broadly summarize
the results, all panels show a similar range of the proportion of
systematic variance attributable to intrapersonal change. Across panels,
the median proportion of \(\tau\) is 0.094 (mean of 0.098), with an
interquartile range of 0.053 to 0.139. All panels include questions
where \(\tau\) is essentially 0, and all panels include questions with
\(\tau\) values greater than 0.20. The maximum \(\tau\) value is 0.26.

\begin{center}
<about here>
\end{center}

Figure 1: Proportions of explained variance in items of personal
culture; based on predictions at participants' wave mid-point.

@@

To the extent that there are differences across the panels, the PSID has
the highest \(\tau\) values with mean .120 and median .135. While we
cannot disentangle features of the sample from features of the questions
asked to each sample, the specific samples for many PSID questions have
lower average ages than those from other panels. To the extent that
younger respondents might be more likely to make durable changes of
opinion, this higher explanatory power of intrapersonal change might
reflect the distinct age profile of respondents in this sample. At the
other end, the GSS has the lowest range of \(\tau\) with mean .073 and
median .069. This potentially reflects the fact that the GSS observes
people for a shorter duration, on average, than the other panels. If,
consistent with life course adaption theories, people are more likely to
make significant cultural changes the longer we observe them, then
duration likely affects the range of \(\tau\) (a point we explore more
below). However, the GSS results are still consistent with results from
the other panels.

While there are some differences between panels, these differences are
small compared to the differences within panels. For about 6 percent of
items explored here, \(\tau\) is greater than 0.20. These questions tend
to ask about external referents (e.g., confidence in specific government
leaders or political parties), life satisfaction, or current financial
position. At the other end, questions about religious identification,
views on gender roles, and support for civil liberties tend to have very
low proportion of variance attributable to intrapersonal change.

@@

Appendix A presents the estimated proportion of variance attributable to
interpersonal differences and intrapersonal change, the estimated values
of \(\tau\), and the proportion of residual variance for each question
analyzed. In general, interpersonal differences are almost always the
largest component of the total variance and tend to explain between 55
and 70 percent of total variance, while intrapersonal change is always
the smallest, typically explaining between 3 and 8 percent of total
variance. Residual variance tends to account for between 22 and 37
percent of variance, though on several questions residual variance is
greater than 50 percent.

@@

Figure 2 plots the results from an ordinary least squares regression of
\(\tau\) as a function of question, panel, and sample features. These
models also include fixed effects for panels, so coefficients reflect
the association between a particular sample or question characteristic
within a panel.{[}KK NOTE: CAN WE SUMMARIZE THESE PANEL FIXED EFFECTS AT
ALL? AE: WE COULD SHOW THAT THEY DON'T CAPTURE TOPIC EFFECTS BEYOND OUR
CLASSIFICATION IF WE INCLUDE THOSE / THAT THEY POSSIBLY CAPTURE
UNMEASURED COUNTRY/CONTEXT DIFFERENCES OR DIFFERENCES IN SURVEY
METHODOLOGY?{]}

\begin{center}
<about here>
\end{center}

Figure 2: Proportion of explained variance attributable to intrapersonal
change; based on predictions at participants' wave mid-point; supressed
intercept model; survey indicators not shown.

Figure 2 shows that the more response options respondents are given and
the larger the sample, the larger the values of \(\tau\). We interpret
these coefficients as suggesting greater resolution on a question makes
it easier to detect and model change. The earlier a question was asked,
even net of how long it was been asked, the lower the \(\tau\).

The most notable result in the meta analysis is the negative coefficient
attached to the duration of years covered by the question. The longer a
question is observed, the less is the value of \(\tau\). Theories of
personal cultural change that link changes in personal culture to social
experiences, including the LCAM, suggest that the longer we observe
respondents, the more likely people are to undergo potentially
transformative experiences and therefore the more variance would be
explained by intrapersonal change. The fact that we find a negative
coefficient attached to this variable can be a challenge to that
assumption {[}TK Note: since there is a complication of this, I changed
is to can be, but feel free to ignore it.{]}

As a further test of this coefficient, we compared the \(\tau\) values
when using the full duration of a panel compared to when we dropped the
final wave and therefore reduced the total duration of observation for
the question. If the coefficient reflects a true negative effect of
duration on \(\tau\), we should see that same effect within questions.
Results from this analysis are presented in Appendix B and contradict
the coefficient from the regression model: we found that the longer we
observe the same question, the higher the \(\tau\) is. We interpret this
combination of findings as suggesting that the kinds of questions asked
for a longer time period tend to demonstrate less intrapersonal change
than questions asked over shorter periods, rather than a true function
of time.

This reinforces the explanation for why the GSS displays the lowest
average proportion of systematic variance attributable to intrapersonal
change: its shorter duration.

While we did not generate expectations for how question content relates
to \(\tau\), some of the associations between question structure, panel
duration, and panels themselves might be driven by differences in the
topics addressed by each panel. To address this, we coded each question
as falling into one of nine different content categories and included
these as indicator variables in the meta analysis. Appendix C presents
the estimates for these content indicators. They show some associations
with \(\tau\), with questions about subjective SES; social life, social
cohesion, and trust; environment and climate; and health and morale
showing larger \(\tau\) values than questions about religion and
spirituality; politics, government and the economy; and gender and
family. {[}KKNOTE: I FEEL LIKE WE WANTED TO SAY SOMETHING ELSE HERE BUT
I CANT REMEMBER WHAT{]}

@@

To test these competing propositions, we use the models presented above
to separately calculate \(\tau\) values for people with at least a
bachelor's degree and people with less than a bachelor's degree at wave
1\footnote{A small number of respondents report different highest
  degrees at wave 1, wave 2, and wave 3. Some of this is due to
  measurement error, and some of it is due to a small number of people
  obtaining a higher degree during the four years of the panel.
  Estimating the panel with highest at wave 1 or highest degree reported
  across the panel produces functionally identical results.} of the
General Social Survey's panels.\footnote{We estimate the models
  presented in Figures 1 and 2 using the full sample, but only use
  respondents who have information on their highest degree to generate
  the results presented in Figures 3 and 4. In total, \#\#\# respondents
  (\%\%\% percent) reported no degree information.} We focus on the GSS
because it contains the largest number of questions tapping general
political dispositions, which is the domain where education has proven
particularly relevant for understanding attitude stability. The GSS also
covers a turbulent window of American politics from 2006 to 2014. This
window covers the start of the Great Recession, debates about federal
intervention in and regulation of Wall Street, the election of Barack
Obama as the first black president in U.S. history, debates about the
role of the federal government in the health care sector, the emergence
of the Tea Party, and political realignment and clarification on the
issue of gay marriage, among other topics.

Figure 3 plots the distribution of differences in \(\tau\) values
between people with at least a bachelor's degree and people with less
than a bachelor's degree at wave 1 of the panel for all 183 GSS items we
explored. Values greater than 0 indicate that intrapersonal change is
more important in explaining the responses of college graduates than
those of people without a college degree, while values less than 0
indicate that intrapersonal change is less important for explaining
variation among people with a college degree than those without a
college degree.

\begin{center}
<about here>
\end{center}

Figure 3: Difference between share of intrapersonal change and share of
inter-personal differences in explained variance; based on predictions
at participants' wave mid-point

@@

To more clearly illustrate some of these differences, we highlight eight
questions designed to tap general political dispositions: partisan
identification (Democrat vs.~Republican) and ideological identification
(liberal vs.~conservative) on seven-point scales; four questions about
the government's role in improving the condition of the poor, paying
people's medical bills, giving special treatment to black people, and
doing things that private businesses could do, measured on five-point
scales; a question about whether the government should do more to reduce
income differences, measured on a seven-point scale; and one question
about whether black people should be given preferences in hiring,
measured on a five-point scale. Estimates of \(\tau\) for these eight
questions, for both education groups, are presented in Figure 4.

\begin{center}
<about here>
\end{center}

Figure 4: Proportions of intrapersonal change of explained variance by
college degree; based on predictions at participants' wave mid-point.

On all eight questions presented in Figure 4, \(\tau\) values are
greater for people without a college degree than for people with a
college degree, meaning intrapersonal change explains less systematic
variance for people with a college degree than for people without one.
There are also large differences in \(\tau\) values across questions for
both groups. For example, on the question of whether the government
should try to solve more problems or leave those problems to be solved
by private businesses (``government do more or less''), less than one
percent of the systematic variance is attributable to intrapersonal
change for both groups. In other words, while people might vacillate on
this question at random (37 percent of variance is residual for this
question), there is functionally no evidence that people make systematic
changes of opinion on this issue during the GSS panel.

In contrast, partisan identification and political ideology both show
larger values of \(\tau\) than most other questions, as well as a larger
absolute difference between people with a college degree and those
without. Compared to the other questions explored here, change as adults
plays a much larger role in explaining why people report the partisan
identification and political ideology they do. And this is particularly
true for respondents who do not have a college degree. The \(\tau\)
value for ideological identification among non-college educated
respondents is almost four times that of college-educated respondents.

>>>>

Does personal culture --an individual's attitudes, beliefs, values, and
practices--- change over the life course, or is it largely fixed once a
person reaches adulthood?

This question underlies several contemporary debates in sociology (Kiley
and Vaisey 2020; Lersch 2023; Lizardo 2017). For example, pragmatist
theories of action claim that changes in social environments cause
people to adapt their views and make new cultural meanings (Gross 2009;
Swidler 2001), while Bourdieusian practice theories argue that the
``past conditions of production'' leave a mark on people's personal
culture that lasts throughout their lives (Bourdieu 1990). Similarly,
models of social influence assume that people adapt their culture in the
face of new information, while the emphasis on cohort effects in models
of aggregate social and cultural change requires them to be open to
change while young but become fairly resistant to it as they age (Ryder
1965). Finally, life course theories posit slower but still important
changes over time as people advance through important transitions in
their lives (Bardi et al. 2009; Elder, Johnson, and Crosnoe 2003).

Because all of the above theoretical perspectives have some empirical
support, the current debate is not about whether they exist, but about
their contribution to explaining cultural differences. That said,
scholars have struggled to reach a consensus on the relative importance
of change during adulthood. Over the span of a few years, most apparent
changes in people's responses to personal survey items appear to be
transitory, with little evidence of persistent change among adults
(Kiley and Vaisey 2020; Vaisey and Kiley 2021). This suggests that
change during adulthood is not a major factor in explaining personal
cultural differences. When looked at over longer time horizons, however,
there is at least some evidence that adults make persistent cultural
changes (Lersch 2023).

These seemingly inconsistent findings are partly due to the fact that
researchers have taken a ``tournament of models'' approach in
adjudicating different theoretical perspectives (Lersch 2023: p.~228).
That is, researchers have compared models of responses to cultural items
over time as consistent with either ``change'' or ``no change,'' which
essentially asks whether we observe \emph{any} intrapersonal change in
personal culture.

We argue that asking only \emph{whether} people change obscures what are
actually highly compatible empirical patterns. Being able to detect
change in a given dataset depends not only on sample populations, item
measurement, statistical methods, and power, but also on definitions of
what counts as ``change.'' These methodological decisions potentially
limit the possibility of convergence on the theoretical debates.
Moreover, asking whether there is evidence of change limits the kind of
questions that can be asked to, ``Do we see change in this measure?''
But such inquiries are ill-suited to address the core of the theoretical
debate: determining \emph{how much} the process of intrapersonal change
contributes to cultural differences.

There is enough evidence to assert that neither pre-adult socialization
nor the contemporary environment alone can fully explain cultural
differences. The question that remains is about the relative
contributions of these processes. In the current debate, for a single
construct (e.g., support for gay rights), existing approaches can only
make claims about \emph{whether} intrapersonal change is happening, not
about how much. This limitation also prevents investigating under what
conditions, across which groups, and in what domains do we see
differences in the relative importance of intrapersonal change. These
questions are about \emph{degree}, not existence, and thus require
precise measurement rather than declarations of victory for a particular
perspective.

In this paper, we intervene in this debate by moving beyond the
``tournament of models'' approach toward a quantification of the
processes that lead to differences in personal culture, and introducing
a method for quantifying the relative contributions of interpersonal
differences and intrapersonal change to the patterns we see in empirical
data. Drawing on the seven panel datasets analyzed in the debate, we use
established methods to quantify the amount of variance in responses that
is due to between-people differences at a single point during the survey
versus the amount of variance in responses that is due to linear
intrapersonal change over time. This measure gives us a sense of the
importance of change over the duration of the panel relative to the
differences that exist between people assuming they never changed.

@@

Despite differences in how these researchers interpret their findings,
the empirical results are not far apart. Kiley and Vaisey (2020) found
that the majority of items they tested favored the AUM, meaning there
was evidence of durable change among adults, a pattern consistent with
Lersch's results. While Lersch (2023) found evidence of durable change
on most items, this change only accounted for a small proportion of
variance explained. On average, people changed only about .07 standard
deviations over 10 years. This is consistent with Kiley and Vaisey's
finding that durable change is often small. And a quarter of the items
Lersch studied still preferred the SDM, suggesting that even with
different assumptions assumptions about change and more extensive data,
many questions show no evidence of durable change. Likewise, in studies
showing evidence of durable change in specific measures of personal
culture, the change is often small. For example, Brocic and Miles (2021)
estimate that completing graduate degrees in humanities, arts, and
social sciences shift peoples' moral relativism only about .2 standard
deviations on average compared to people with no college degree, and
this is the largest effect they identify. And studies of aggregate
change show that, even on items where cohort effects explain more
variance than period effects, there is always evidence that some people
change over time (Vaisey and Lizardo 2016). In other words, despite
different interpretations, the results of previous work are in many ways
highly consonant.

@@

A more theoretically productive approach begins with a model assuming
that during an observed time period people might remain stable, might
change a little, or might undergo significant shifts in their personal
culture. Lersch's LCAM formally does this by modeling each individual as
following their own trajectory over time. But rather than debating
whether this model fits a particular data set better than a model that
assumes that people never change, researchers should quantify individual
changes over time and compare those to the differences that exist
between people at the panel's start. In other words, the focus should
shift from asking, ``do people change?'' to asking, ``what are the
relative contributions of change and initial differences toward
explaining differences in personal culture?''

In this approach, there are two metrics that each reflect a combination
of relevant theoretical processes. The first metric, interpersonal
differences at the start of the survey, reflects the accumulated
experiences of people prior to entering the panel survey that shape how
they answer questions, what Lersch calls ``early imprinting.'' While
commonly associated with experiences during an early formative period
that result in settled dispositions, this variation could also reflect
experiences that happen at any stage in life as long as they predate the
panel and consistently impact subsequent responses. For instance, for
those people entering the panel post-retirement, this ``imprinting''
might reflect this pivotal life transition. Consequently, this metric
also reflects variation in individuals' social roles or statuses at the
start of the panel that were important in shaping their dispositions.

@@

A third metric, often called ``residual variance'' or ``fluctuation,''
accounts for the remainder of variance in peoples' responses over time.
This non-durable change emerges for a variety of reasons. For example,
people might not have clear personal cultural dispositions as measured
by surveys. Instead, they might internalize a broad set of
considerations and construct an opinion in the context of the survey
interview, with different considerations coming to the forefront of
their cognition each interview (Feldman and Zaller 1992; Tourangeau,
Rips, and Rasinski 2000; Zaller 1992). This variance in considerations
can produce variance in responses even if peoples' ``dispositions''
--the full set of considerations they could draw on-- remains stable
over time. Additionally, this variance can also include measurement
errors, such as misinterpreted questions, erroneous response selections,
or responses getting coded incorrectly. While this third process
reflects important aspects of personal culture, it does not directly
touch on the ongoing debate outlined here. Because of this we quantify
this component but focus principally on the other two.

@@

Third, we expect the variance explained by each process
-\/--interpersonal differences and intrapersonal change-\/-- to depend
on specific attributes of the question and panel. First, the higher the
response resolution for a given question and the larger the sample, the
greater the variance explained by intrapersonal change should be, as
smaller changes in intrapersonal change will be easier to detect.
Second, the LCAM and various theoretical perspectives that underlie it
posit that social transitions drive changes in personal culture. Since
the probability that any person experiences such a transition goes up
the longer we observe them, we expect panel duration to be positively
related to the proportion of variation explained by intrapersonal
change.

@@

\[
y_{i, t} = \beta_0 + u_{0, i} + \beta_{1} x_t + u_{1, i} x_{i, t} + \epsilon_{i, t}
\] \[
u_{0,i} \sim \mathcal{N}(0,\sigma_{\mu_0}^2) \\
\] \[
u_{1,i} \sim \mathcal{N}(0,\sigma_{u_1}^2)
\]

@@

\[
\omega = \frac{V(C)}{V(C) + V(D)}
\]

We can think of this number, \(\omega\) as akin to an Intraclass
Correlation Coefficient focused only on the systematic components of the
model while ignoring residual variance that might be attributable to
measurement error or non-persisting change (see also the variance
decomposition of period and cohort by Vaisey and Lizardo 2016). The
higher this number is, the more systematic variance in responses is
attributable to intrapersonal change during the panel. The lower the
number, the less intrapersonal change seems to play a large role in
explaining systematic cultural differences during the panel. Moving
forward, we use \(\omega\) as a placeholder for \emph{systematic
variance attributable to intrapersonal change}.

@@

As a second step in our analysis, we use ordinary least squares
regression to model \(\omega\) as a function of features of the question
and panel. This allows us to explore how these features affect the
prevalence of intrapersonal change. These covariates include the number
of response options (measured as 2, 3-5, and more than 5), the panel,
the log of the number of participants, the number of waves observed, the
time since the question was first asked, and the total span of time the
question was asked.

@@

Figure 1 plots the proportion of systematic variance attributable to
interpersonal differences and intrapersonal change for each of the 609
questions analyzed, plotted separately by panel. To broadly summarize
the results, all panels show a similar range of the proportion of
systematic variance attributable to intrapersonal change. Across panels,
the median proportion of \(\omega\) is 0.094 (mean of 0.098), with an
interquartile range of 0.053 to 0.139. All panels include questions
where \(\omega\) is essentially 0, and all panels include questions with
\(\omega\) values greater than 0.20. The maximum \(\omega\) value is
0.26.

\begin{center}
<about here>
Figure 1: Proportions of Systematic Variance in Personal Culture
\end{center}

@@

To the extent that there are differences across the panels, the PSID has
the highest \(\omega\) values with mean .120 and median .135. While we
cannot disentangle features of the sample from features of the questions
asked to each sample, the specific samples for many PSID questions have
lower average ages than those from other panels. To the extent that
younger respondents might be more likely to make durable changes of
opinion, this higher explanatory power of intrapersonal change might
reflect the distinct age profile of respondents in this sample. At the
other end, the GSS has the lowest range of \(\omega\) with mean .073 and
median .069. This potentially reflects the fact that the GSS observes
people for a shorter duration, on average, than the other panels. If,
consistent with life course adaption theories, people are more likely to
make significant cultural changes the longer we observe them, then
duration likely affects the range of \(\omega\) (a point we explore more
below). However, the GSS results are still consistent with results from
the other panels.

While there are some differences between panels, these differences are
small compared to the differences within panels. For about 6 percent of
items explored here, \(\omega\) is greater than 0.20. These questions
tend to ask about external referents (e.g., confidence in specific
government leaders or political parties), life satisfaction, or current
financial position. At the other end, questions about religious
identification, views on gender roles, and support for civil liberties
tend to have very low proportion of variance attributable to
intrapersonal change.

@@

Supplemental Materials A presents the estimated proportion of variance
attributable to interpersonal differences and intrapersonal change, the
estimated values of \(\omega\), and the proportion of residual variance
for each question analyzed. In general, interpersonal differences are
almost always the largest component of the total variance and tend to
explain between 55 and 70 percent of total variance, while intrapersonal
change is always the smallest, typically explaining between 3 and 8
percent of total variance. Residual variance tends to account for
between 22 and 37 percent of variance, though on several questions
residual variance is greater than 50 percent.

@@

Figure 2 plots the results from an ordinary least squares regression of
\(\omega\) as a function of question, panel, and sample features. These
models also include fixed effects for panels and topics, so coefficients
reflect the association within a panel and topical domain.

\begin{center}
<about here>
Figure 2: Regression Model Estimating $\omega$
\end{center}

Figure 2 shows that the more response options respondents are given and
the larger the sample, the larger the values of \(\omega\). We interpret
these coefficients as suggesting greater resolution on a question makes
it easier to detect and model change. The earlier a question was asked,
even net of how long it was been asked, the lower the \(\omega\).

While we did not generate expectations for how question content relates
to \(\omega\), some of the associations between question structure,
panel duration, and panels themselves might be driven by differences in
the topics addressed by each panel. To address this, we coded each
question as falling into one of nine different content categories and
included these as indicator variables in the meta analysis. Supplemental
Materials B presents the estimates for these content indicators. They
show some associations with \(\omega\), with questions about subjective
SES; social life, social cohesion, and trust; environment and climate;
and health and morale showing larger \(\omega\) values than questions
about religion and spirituality; politics, government and the economy;
and gender and family.

The most notable result in the meta analysis is the negative coefficient
attached to the duration of years covered by the question. The longer a
question is observed, the less is the value of \(\omega\). Theories of
personal cultural change that link changes in personal culture to social
experiences, including the LCAM, suggest that the longer we observe
respondents, the more likely people are to undergo potentially
transformative experiences and therefore the more variance would be
explained by intrapersonal change. The fact that we find a negative
coefficient attached to this variable can be a challenge to that
assumption.

As a further test of this coefficient, we compared the \(\omega\) values
when using the full duration of a panel compared to when we dropped the
final wave and therefore reduced the total duration of observation for
the question. If the coefficient reflects a true negative effect of
duration on \(\omega\), we should see that same effect within questions.
Results from this analysis are presented in Supplemental Materials C and
contradict the coefficient from the regression model: we found that the
longer we observe the same question, the higher the \(\omega\) is. We
interpret this combination of findings as suggesting that the kinds of
questions asked for a longer time period tend to demonstrate less
intrapersonal change than questions asked over shorter periods, rather
than a true function of time. This reinforces the explanation for why
the GSS displays the lowest average proportion of systematic variance
attributable to intrapersonal change: its shorter duration.

@@

To test these competing propositions, we use the models presented above
to separately calculate \(\omega\) values for people with at least a
bachelor's degree and people with less than a bachelor's degree at wave
1\footnote{A small number of respondents report different highest
  degrees at wave 1, wave 2, and wave 3. Some of this is due to
  measurement error, and some of it is due to a small number of people
  obtaining a higher degree during the four years of the panel.
  Estimating the panel with highest at wave 1 or highest degree reported
  across the panel produces functionally identical results.} of the
General Social Survey's panels. We focus on the GSS because it contains
the largest number of questions tapping general political dispositions,
which is the domain where education has proven particularly relevant for
understanding attitude stability. The GSS also covers a turbulent window
of American politics from 2006 to 2014. This window covers the start of
the Great Recession, debates about federal intervention in and
regulation of Wall Street, the election of Barack Obama as the first
black president in U.S. history, debates about the role of the federal
government in the health care sector, the emergence of the Tea Party,
and political realignment and clarification on the issue of gay
marriage, among other topics.

Figure 3 plots the distribution of differences in \(\omega\) values
between people with at least a bachelor's degree and people with less
than a bachelor's degree at wave 1 of the panel for all 183 GSS items we
explored. Values greater than 0 indicate that intrapersonal change is
more important in explaining the responses of college graduates than
those of people without a college degree, while values less than 0
indicate that intrapersonal change is less important for explaining
variation among people with a college degree than those without a
college degree.

\begin{center}
<about here>
Figure 3: Difference in $\omega$ Across College Graduates and Non-Graduates
\end{center}

@@

To more clearly illustrate some of these differences, we highlight eight
questions designed to tap general political dispositions: partisan
identification (Democrat vs.~Republican) and ideological identification
(liberal vs.~conservative) on seven-point scales; four questions about
the government's role in improving the condition of the poor, paying
people's medical bills, giving special treatment to black people, and
doing things that private businesses could do, measured on five-point
scales; a question about whether the government should do more to reduce
income differences, measured on a seven-point scale; and one question
about whether black people should be given preferences in hiring,
measured on a five-point scale. Estimates of \(\omega\) for these eight
questions, for both education groups, are presented in Figure 4.

\begin{center}
<about here>
Figure 4: Difference in $\omega$ Across College Graduates and Non-Graduates on Political Culture
\end{center}

On all eight questions presented in Figure 4, \(\omega\) values are
greater for people without a college degree than for people with a
college degree, meaning intrapersonal change explains less systematic
variance for people with a college degree than for people without one.
There are also large differences in \(\omega\) values across questions
for both groups. For example, on the question of whether the government
should try to solve more problems or leave those problems to be solved
by private businesses (``government do more or less''), less than one
percent of the systematic variance is attributable to intrapersonal
change for both groups. In other words, while people might vacillate on
this question at random (37 percent of variance is residual for this
question), there is functionally no evidence that people make systematic
changes of opinion on this issue during the GSS panel.

In contrast, partisan identification and political ideology both show
larger values of \(\omega\) than most other questions, as well as a
larger absolute difference between people with a college degree and
those without. Compared to the other questions explored here, change as
adults plays a much larger role in explaining why people report the
partisan identification and political ideology they do. And this is
particularly true for respondents who do not have a college degree. The
\(\omega\) value for ideological identification among non-college
educated respondents is almost four times that of college-educated
respondents.

We note here that this meaningful difference in \(\omega\) values across
education groups and across questions would not be detectable using
previous methods. For partisan identification and political ideology,
both college-educated respondents and people with less than college
degree would likely favor the AUM or LCAM over the SDM because they
these questions both show evidence of some members of the population
making intrapersonal change. In other words, using the tournament of
models approach obscures the fact that the relative importance of
intrapersonal change in explaining differences on these two items is
different across these two groups, that intrapersonal change explains
more systematic variance for ideological identification than partisan
identification for people without a college degree, and that there
appears to be more durable change on questions of affirmative action
than on questions of government aid to black Americans.
